---
author: "Hector Roux de Bézieux"
date: '`r format(Sys.time(), "%d %B , %Y")`'
title: "Diagram for Dune"
output:
  html_document:
    toc: TRUE
    toc_depth: 2
    toc_float: TRUE
    number_sections: TRUE
    code_download: TRUE
---

We use a subset of the Allen Smart-Seq nuclei dataset. Run `?Dune::nuclei` for more details on pre-processing.

```{r setup}
suppressPackageStartupMessages({
  library(RColorBrewer)
  library(dplyr)
  library(ggplot2)
  library(tidyr)
  library(knitr)
  library(purrr)
  library(Dune)
  library(here)
  library(DailyHRB)
  library(cowplot)
})
data("nuclei", package = "Dune")
```

# Panel a
## Initial 2D visualization

We can also see how the three clustering algorithm partitioned the dataset initially:

```{r}
nuclei <- nuclei %>% filter(!SC3 %in% c(4, 24, 28),
                            Monocle != 23,
                            !Seurat %in% c(2, 8))
inits <- map(c("SC3", "Seurat", "Monocle"), function(clus_algo){
  df <- nuclei
  df$clus_algo <- nuclei[, clus_algo]
  cols <- brewer.pal(n = n_distinct(df$clus_algo), name = "Dark2")
  names(cols) <- unique(df$clus_algo)

  p <- ggplot(df, aes(x = x, y = y, col = as.character(clus_algo))) +
    geom_point(size = 2, alpha = .8) +
    guides(color = FALSE) +
    ggtitle(label = clus_algo) +
    scale_color_manual(values = cols) +
    theme_bw() + 
    theme(plot.title = element_text(hjust = .5, size = 25),
          rect = element_blank(),
          panel.border = element_blank(),
          axis.title = element_blank(),
          axis.text = element_blank(),
          axis.ticks = element_blank())
  return(p)
})
```

## ARI matrix

The adjusted Rand Index between the three methods can be computed.

```{r}
ARI_mat <- ARIs(nuclei %>% select(SC3, Seurat, Monocle)) %>%
  as.data.frame() %>%
  mutate(method1 = rownames(.)) %>%
  pivot_longer(-method1, names_to = "method2", values_to = "ARI") %>%
  filter(method1 != method2) %>%
  group_by(method1, method2) %>%
  mutate(pair = paste0(max(method1, method2), " - ", min(method1, method2))) %>%
  ungroup() %>%
  select(pair, ARI) %>%
  distinct() %>%
  arrange(pair)
ARI <- ggplot(ARI_mat, aes(x = pair, y = 0)) +
  geom_tile(aes(fill = ARI), col = "black", size = 1) +
  scale_fill_gradientn(colours = brewer.pal(9, "Spectral"),
                         limits = c(0, 1)) +
  geom_text(aes(label = round(ARI, 2)), size = 8) +
  my_theme() +
  guides(fill = FALSE) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_text(size = 8, colour = "black")) +
  scale_x_discrete(
    limits = c("Seurat - SC3", "SC3 - Monocle", "Seurat - Monocle"))
rm(ARI_mat)
```

## Pairwise confusion matrix

As we can see, the ARI between the three methods is initially quite low.

```{r}
pairs <- combn(c("SC3", "Seurat", "Monocle"), 2) %>%
  as.data.frame(stringsAsFactors = FALSE)
confusion <- lapply(pairs, function(pair){
  p <- ConfusionPlot(nuclei[, pair[1]], nuclei[, pair[2]]) +
    labs(x = pair[1], y = pair[2]) +
    scale_color_gradientn(colours = brewer.pal(11, "Spectral"),
                          limits = c(0, 1)) +
    scale_size_continuous(limits = c(1, 420)) +
    guides(col = FALSE, size = FALSE)
  return(p)
})
rm(pairs)
```

```{r}
legend_size <- get_legend(
  confusion[[1]] +
    theme(legend.position = "right") +
    guides(size = guide_legend(title.position = "top", fill = "grey",
                               title.hjust = .5)) +
    theme(legend.margin = margin(0, 0, 0, 0),
          legend.text = element_text(size = 5),
          ))
legend_colour <- get_legend(
  confusion[[1]] +
    theme(legend.position = "right") +
    guides(col = guide_colourbar(title.position = "top", title.hjust = .5)) +
    theme(legend.margin = margin(0, 0, 0, 0),
          legend.text = element_text(size = 5),
          ))
legend <- plot_grid(legend_size, legend_colour, ncol = 2)
rm(legend_colour, legend_size)
```

## Full panel

```{r}
first_row <- plot_grid(plotlist = inits, ncol = 3, scale = .95)
second_row <- plot_grid(plotlist = confusion, ncol = 3, scale = .95)
ARI <- plot_grid(NULL, ARI, NULL, ncol = 1,
                 rel_heights = c(.4, 1, .4))
third_row <- plot_grid(legend, ARI, NULL, ncol = 3, rel_widths = c(1, 1, 1), 
                       scale = .95)
p_a <- plot_grid(first_row, NULL, second_row, NULL, third_row,
                 ncol = 1, rel_heights = c(2, .5, 2, .3, 1))
rm(first_row, second_row, third_row, inits, confusion, legend, ARI)
```

# Impact of merging
## Confusion matrix with rectangle

```{r}
pOG <- ConfusionPlot(nuclei$SC3, nuclei$Monocle) +
  labs(x = "SC3", y = "Monocle") +
  scale_color_gradientn(colours = brewer.pal(11, "Spectral"),
                          limits = c(0, 1)) +
    scale_size_continuous(limits = c(1, 420)) +
  guides(col = FALSE, size = FALSE) +
  annotate("tile",  x = 4.5, y = "20", fill = "NA", col = "black", width = 2,
           height = 5, size = 2) +
  annotate("curve", x = 4, y  = "20", yend = "20", xend = 5,
           arrow = arrow(length = unit(5, "pt"))) +
  annotate("curve", x = 4, y  = "19", yend = "19", xend = 5,
           arrow = arrow(length = unit(5, "pt")), curvature = -.5)
pOG
```

## Confusion matrix after merging

```{r}
df <- nuclei
SC3 <- nuclei$SC3
SC3[SC3 %in% c(21, 20)] <- 20
df$SC3 <- SC3
pMerg <- ConfusionPlot(df$SC3, df$Monocle) +
  labs(x = "SC3", y = "Monocle") +
  scale_x_discrete(limits = levels(pOG$data$x)) +
  scale_y_discrete(limits = levels(pOG$data$y)) +
  scale_color_gradientn(colours = brewer.pal(11, "Spectral"),
                          limits = c(0, 1)) +
  scale_size_continuous(limits = c(1, 420)) +
  guides(col = FALSE, size = FALSE) +
  annotate("tile",  x = 4.5, y = "20", fill = "NA", col = "black", width = 2,
           height = 5, size = 2)

pMerg
rm(SC3, nuclei)
```

## ARI matrix

```{r}
ARI_mat <- ARIs(df %>% select(SC3, Seurat, Monocle)) %>%
  as.data.frame() %>%
  mutate(method1 = rownames(.)) %>%
  pivot_longer(-method1, names_to = "method2", values_to = "ARI") %>%
  filter(method1 != method2) %>%
  group_by(method1, method2) %>%
  mutate(pair = paste0(max(method1, method2), " - ", min(method1, method2))) %>%
  ungroup() %>%
  select(pair, ARI) %>%
  distinct() %>%
  arrange(pair)
ARI_Final <- ggplot(ARI_mat, aes(x = pair, y = 0)) +
  geom_tile(aes(fill = ARI), col = "black", size = 1) +
  scale_fill_gradientn(colours = brewer.pal(9, "Spectral"),
                         limits = c(0, 1)) +
  geom_text(aes(label = round(ARI, 2)), size = 8) +
  my_theme() +
  guides(fill = FALSE) +
  theme(axis.title = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_text(size = 8, colour = "black")) +
  scale_x_discrete(
    limits = c("Seurat - SC3", "SC3 - Monocle", "Seurat - Monocle"))
rm(ARI_mat, df)
```

## Full panel

```{r}
ARI_Final <- plot_grid(NULL, ARI_Final, NULL, ncol = 1, 
                       rel_heights = c(1, 1, 1)) 
p_b <- plot_grid(pOG, pMerg, NULL, ARI_Final, ncol = 4, 
                 rel_widths = c(1, 1, .2, 1), scale = .95)
rm(ARI_Final, pMerg, pOG)
```

# All together

```{r}
p <- plot_grid(p_a, NULL, p_b, rel_heights = c(10, .5, 3), ncol = 1,
               labels = c("a)", "", "b)"), label_size = 30)
save_plot(filename = here("Figure-Diagram", "figures", "panel.svg"),
          plot = p, base_height = 13.5, base_width = 10)
```