---
author: "Hector Roux de Bézieux"
date: '`r format(Sys.time(), "%d %B , %Y")`'
title: "Dune's inner workings"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: TRUE
    number_sections: true
    code_download: TRUE
---

We use a subset of the Allen Smart-Seq nuclei dataset. Run `?Dune::nuclei` for more details on pre-processing.

```{r setup}
suppressPackageStartupMessages({
  library(RColorBrewer)
  library(dplyr)
  library(ggplot2)
  library(tidyr)
  library(knitr)
  library(purrr)
  library(Dune)
  library(here)
})
data("nuclei", package = "Dune")
theme_set(theme_classic())
```

# Initial visualization

We can also see how the three clustering algorithm partitioned the dataset initially:

```{r}
walk(c("SC3", "Seurat", "Monocle"), function(clus_algo){
  df <- nuclei
  df$clus_algo <- nuclei[, clus_algo]
  p <- ggplot(df, aes(x = x, y = y, col = as.character(clus_algo))) +
    geom_point(size = .5) +
    guides(color = FALSE) +
    labs(title = clus_algo)
  ggsave(here("Figure-Diagram", "figures", paste0("Init-", clus_algo, ".pdf")),
         p, height = 3, width = 5)
})
```

# Merging with **Dune**
## Initial ARI

The adjusted Rand Index between the three methods can be computed.

```{r}
ggsave(filename = here("Figure-Diagram", "figures", "ARI_init.pdf"),
       height = 7, width = 7,
       plot = plotARIs(nuclei %>% select(SC3, Seurat, Monocle)))
       
```

As we can see, the ARI between the three methods is initially quite low.

```{r}
walk(combn(c("SC3", "Seurat", "Monocle"), 2) %>%
       as.data.frame(stringsAsFactors = FALSE), 
     function(pair){
      p <- ConfusionPlot(nuclei[, pair[1]], nuclei[, pair[2]]) +
        labs(x = pair[1], y = pair[2])
      ggsave(plot = p, 
             filename = here("Figure-Diagram", "figures",
             paste0("Conf_init-", pair[1], "-", pair[2], ".pdf")),
             width = 5, height = 3)
     })
```


# Merging one cluster

```{r}
Monocle <- nuclei$Monocle
Monocle[Monocle %in% c(3, 16)] <- 3
nuclei$Monocle <- Monocle
```

## Actual merging

We can now try to merge clusters with the `Dune` function. At each step, the algorithm will print which clustering label is merged (by its number, so `1~SC3` and so on), as well as the pair of clusters that get merged.

```{r}
merger <- Dune(clusMat = nuclei %>% select(SC3, Seurat, Monocle),
               verbose = TRUE, keepAll = TRUE)
```

The output from `Dune` is a list with four compoments:

```{r}
names(merger)
```

`initialMat` is the initial matrix. of cluster labels. `currentMat` is the final matrix of cluster labels. `merges` is a matrix that recapitulates what has been printed above, while `ImpARI` list the ARI improvement over the merges.

## ARI improvement

We can now see how much the ARI has improved:

```{r}
plotARIs(clusMat = merger$currentMat)
```

The methods now look much more similar, as can be expected.

We can also see how the number of clusters got reduced.

```{r}
plotPrePost(merger)
```

For SC3 for example, we can visualize how the clusters got merged:

```{r}
ConfusionPlot(merger$initialMat[, "SC3"], merger$currentMat[, "SC3"]) +
  labs(x = "Before merging", y = "After merging")
```

Finally, the ARIImp function tracks mean ARI improvement as pairs of clusters get merged down.

```{r}
ARItrend(merger)
```

# 
```{r}
evol <- merger$intermediary[[2]]
names(evol) <- colnames(merger$initialMat)
evol <- map2(evol, names(evol), function(ARI, clus){
  clusters <- unique(merger$initialMat[, clus]) %>% as.numeric() %>%
    sort() %>% as.character()
  pairs <- combn(clusters, 2)
  df <- data.frame(ARI_imp = ARI,
                   clus1 = pairs[1,],
                   clus2 = pairs[2, ])
}) %>% bind_rows(.id = "clus") %>%
  mutate(clus1 = as.numeric(clus1),
         clus1 = as.factor(clus1),
         clus2 = as.numeric(clus2),
         clus2 = as.factor(clus2))
ggplot(evol, aes(x = clus1, y = clus2, fill = ARI_imp)) +
  geom_tile(col = "black") +
  facet_wrap(~clus, scales = "free") +
  scale_fill_gradientn(colours = c("#54278F", "#9E9AC8", "#006D2C"),
    values = scales::rescale(c(min(evol$ARI_imp), 10^(-3), max(evol$ARI_imp)))) +
  geom_tile(fill = "transparent", col = "red", size = 3,
            data = evol %>% filter(ARI_imp == max(ARI_imp))) +
  scale_x_discrete(position = "top") +
  labs(x = "")

```

# Session

```{r}
sessionInfo()
```
